\documentclass[a4paper,11pt,twoside,openright]{scrbook}

\usepackage{../jnwthesis}
\usepackage{amsmath}
\usepackage{lipsum}
\usepackage{standalone}
\standalonetrue

\bibliography{/Users/jonwells/Documents/bibtex/Thesis}
\graphicspath{{../figs/}}


\begin{document}

\chapter{Methods}\label{chapter:methods}

\section{Introduction}
The methods provided in this chapter relate to work performed either exclusively
by myself or in direct collaboration with others. Except where strictly
necessary, any details relating to work that I was not closely involved with
have not been included - these can instead be found in the online versions of
the published papers included in appendix \ref{appendix:published}. For the sake
of accuracy I have only made minimal changes to descriptions as they appear in
the published documents. Unless indicated otherwise, most of the sections that
follow should therefore be considered as quotes from the published versions,
edited for clarity. In these cases, I have obtained permission for reuse where
required. Additional code relating to this work can be found at:
https://github.com/jonwells90.

\section{Methods}

\subsection{Chapter \ref*{chapter:operons}: Operon gene order is optimised for ordered assembly of protein complexes}

\subsubsection{Protein structural datasets}
We started with the full set of prokaryotic X-ray and electron microscopy
structures in the PDB on June 12, 2014. We considered all heteromeric pairs of
subunits from the same complex, defined as having at least two different protein
chains of at least 30 residues each and mapping to different UniProt sequences
from a single species. Complexes with known quaternary structure assignment
errors \cite{Levy2007} were excluded. Very large complexes with more than 24
subunits were excluded, because we have not shown that the assembly of these can
be predicted accurately from their structures. Heteromeric subunit pairs were
filtered for redundancy at the level of 50\% sequence identity.

\subsubsection{Mapping subunit pairs to operons}
Operon datasets were downloaded from the DOOR2 database \cite{Mao2014}. Relevant
datasets were identified based on the species and strain of each gene pair.
After converting GI numbers to UniProt accession identifiers in each dataset,
the set of gene pairs was mapped to the operon data. Operons encoding both
members of a pair were added to a reference dictionary, with the locus and
directionality of each gene being used to arrange constituent genes in order of
expression. In rare cases where the copy number of a gene within an operon was
found to be greater than one, the position of the gene in the operon was taken
to be that of the first copy to be encountered, reading in the 5' to 3'
direction. The set was then filtered to remove redundant operons (i.e. identical
operons from similar strains or species). In total, 368 gene pairs (220
adjacent) were mapped to 192 unique operons, with the remaining 711 pairs being
expressed in different transcriptional units. Similarly, we also mapped a set of
2,562 binary protein-protein interactions (IM-22059) \cite{Rajagopala2014} to
the E. coli K-12 W3110 operons for the analysis displayed in figure
\ref{figure:intervening}C.

To assess whether the gene order of a pair was evolutionary conserved, we used
the STRING v9.1 database \cite{Franceschini2013}. For each pair, we manually
assessed, using the STRING online interface, whether all occurrences of a given
gene pair shared the same gene order within their local evolutionary group as
defined in STRING. This is at the level of phylum (e.g. Firmicutes or
Euryarchaeota) or class for proteobacteria. Gene pairs present across only a
very limited evolutionary range (less than three genera) were not considered to
be evolutionarily conserved. Gene pairs associated with evolutionary gene fusion
events were identified as those sharing >40\% sequence identity with a gene pair
with evidence for fusion in STRING, similar to what has been done previously
\cite{Marsh2013}.

\subsubsection{Abundance measurements}
We mapped all protein complex subunits in our dataset against the sequences of
prokaryotic proteins from PaxDB v4.0 \cite{Wang2015}, selecting abundance
measurements from proteins with >90\% sequence identity to a subunit. The
results in figures \ref{figure:operonabundance} and
\ref{suppfigure:operonexpression} only use abundance measurements from
\textit{E. coli}, but the analyses in table \ref{table:abundgeneorder} and
figures \ref{suppfigure:operoncomparison}, \ref{suppfigure:geneorder}, and
\ref{suppfigure:operonexpression} are repeated using combined measurements from
all available prokaryotes and also using protein synthesis rates derived from
ribosomal profiling \cite{Li2014b}.

\subsubsection{Prediction of assembly pathways}
Ordered protein complex assembly pathways were predicted in a manner very
similar to what has been done previously \cite{Marsh2013}. First, the complex is
considered in terms of its constituent subunits and the sizes of the interfaces
that can be formed between any pair of subunits are calculated with AREAIMOL
\cite{Winn2011}. Our model assumes that assembly will proceed via formation of
the largest possible interface. The process is then repeated by calculating all
possible interfaces that could form between subunits and subcomplexes until the
full complex is assembled. To define which of a pair of subunits assembles first
and which assembles later, we consider the first step of assembly that brings
the two subunits together within the same (sub)complex. Whichever subunit was
part of a larger subcomplex prior to this step is defined as assembling first.
For example, in the blue pathway in figure \ref{figure:operonassembly}A, the
blue subunit homodimerizes first and then interacts sequentially with the free
red subunits, so the blue subunit is defined as assembling first. If,
alternatively, the first step of assembly had been a heterodimerization between
the blue and red subunits, then both subunits would be classified as assembling
simultaneously. The source code for predicting assembly pathways from protein
complex structures is available at
https://github.com/marshlab/assembly-prediction.

\clearpage
\subsection{Chapter \ref*{chapter:degradation}: Degradation kinetics of proteins
are explained by assembly of protein complexes}

\subsubsection{Protein structural dataset}
Starting from the entire set of protein structures in the Protein Data Bank on
2016-02-24, we searched for all polypeptide chains with >70\% sequence identity
to a human or mouse gene. For genes that map to multiple chains, we selected a
single chain sorting by sequence identity, then number of unique subunits in the
complex, and then the number of atoms present in the chain. Pairwise interfaces
were calculated between all pairs of subunits using AREAIMOL \cite{Winn2011}.
The normalised assembly order was calculated for all complexes, excluding those
containing nucleic acid chains, by first predicting the (dis)assembly pathway as
previously described using all the pairwise interfaces from each heteromeric
complex \cite{Marsh2013} and implemented in the assembly-prediction package
\cite{Wells2016}. For subunits with multiple copies within a single complex, the
average assembly order of each subunit type was considered. The normalised
assembly order was defined so that the first subunit to assemble has a value of
0, the last has a value of 1, and the average value for all unique subunits in a
complex is equal to 0.5.

\subsubsection{Non-structural dataset}
To complement the analysis of protein complexes of known structure, we also
performed coexpression analyses on the non-redundant `core' set of mammalian
complexes from CORUM \cite{Ruepp2009} (downloaded 2015-10-20). As CORUM
preferentially uses human complexes in its non-redundant set, homologous mouse
versions of each complex were generated by replacing each subunit/gene with its
mouse counterpart, provided sequence identity was at least 70\%. Sequence
identities were calculated by collecting all mouse sequences for which NED/ED
classifications were available and running BLAST on these against all genes in
the CORUM core set. In cases where the identity of a subunit was ambiguous (as
defined by CORUM), the first possible subunit for which homology data were
available was selected.

For further validation of the tendency of NED proteins to be `core' subunits,
processed mass spectrometric data was acquired from the dataset published by
Hein et al. \cite{Hein2015}. Their definition of core stoichiometry signature
was used, specifically those proteins matching the criteria of residing in the
circle with radius: 1 (in $log{10}$ units), centred at: -0.5, 0 (abundance
stoichiometry to interaction stoichiometry, see figure 3B\cite{Hein2015}).

\subsubsection{Coexpression analyses}
Coexpression data were downloaded from COXPRESdb \cite{Okamura2014} (mouse
dataset: Mmu.v13-01.G20959-S31479; human dataset: Hsa.v13-01.G20280-S73083). For
each complex, the mean coexpression of each available subunit was calculated,
using pairwise correlations with all other available subunits in the complex.
Cases where fewer than three unique subunits were present in the complex were
discarded, due to calculations of average coexpression for each protein being
superficially identical.

\subsubsection{eQTL analyses}
Human protein complexes were downloaded from the PDB (2016-11-24) using a
minimum sequence identity of 90\% for all chains in order to exclude structures
such as human-viral complexes. eQTL data for 53 tissues was acquired from GTEx
(v6p release) and then mapped to the human complexes. In order to maximise the
available data, tissue datasets were selected by using whichever tissue
maximised the number of available eQTLs for each complex.

\subsection{Chapter \ref*{chapter:aneuploidy}: Autosomal dosage compensation in
aneuploid cells}
\subsubsection{Protein structural dataset}
A large collection of \textit{S. cerevisiae} complexes was compiled from the PDB
by mapping any structures containing chains that mapped with at least 90\%
sequence identity to yeast proteins. Where overlapping subcomplexes existed,
those with the greatest number of unique subunits were retained. Genes from
these complexes were then mapped to the yeast aneuploidy dataset provided in
Dephoure et al. \cite{Dephoure2014}. Wherever genes mapped to multiple PDB
structures, structures were selected on the basis of highest sequence identity.
Assembly order predictions were calculated using the same assembly-prediction
package as previously described \cite{Wells2016}.


\subsubsection{Aggregation dataset}
To briefly summarise, the experimental protocol for acquiring SILAC aggregation
data is essentially the same as that described in Dephoure et al.
\cite{Dephoure2014}, with the exception being that the whole cell lysate is
centrifuged to extract the aggregated fraction separately. The data from the
aggregate fraction in each experiment are initially normalised (by our
collaborators) by subtracting the average of all $log{2}$ disomic ratios from
single-copy genes (which should theoretically be zero) from each protein in the
aggregate.

Further normalisation is carried out using mean centring. Specifically, each
data point is rescaled such that the mean of all data points in each experiment
(i.e. each disomic yeast strain) is equal to the overall mean disomic ratio of
all experiments combined. This has the effect of reducing the variance
attributable to batch effects - see figure \ref{suppfigure:aneuploidy_aggnorm}.

\subsubsection{Normalised abundance calculations}
Protein abundance data for yeast was acquired from PaxDB v4.0 \cite{Wang2015}
and mapped to each gene in the structural dataset. These values were then
normalised as follows:

\begin{displaymath}
    f(x) = \log_2 x - \log_2 m
\end{displaymath}

Where $m$ is the median abundance of subunits within each complex. This
normalisation procedure allows abundance differences to be measured in terms of
$log_{2}$ fold-change.

\subsubsection{Disorder predictions}
Disorder predictions for the complete set of protein sequences from \textit{S.
cerevisiae} were generated using the command line version of IUPred Version 1.0
\cite{Dosztanyi2005a,Dosztanyi2005}, with the output being a disorder score for
each residue in the sequences. An overall score for each protein was given my
taking the mean disorder score across all residues.

\subsection{Chapter \ref*{chapter:hawks}: Hawk proteins: A paralogous family of
eukaryotic SMC-kleisin regulators}\label{subsection:hawkmethods}

\subsubsection{Construction of homology networks}
Proteome fasta files for \textit{S. cerevisiae}, \textit{S. pombe} and
\textit{H. sapiens} were downloaded from the UniProt reference proteomes
databank \cite{Consortium2017} (2016-04) and HHSuite v.3.0.0 was compiled from
source \cite{Soding2005,Remmert2011}. HHsuite databases were constructed as per
the protocol described in the HH-suite manual (available at
http://www.mpibpc.mpg.de/soeding or https://github.com/soedinglab/hh-suite),
using the clustered uniprot20\_2016\_02 database. It should be noted that due to
the fact that HHsuite databases are generated from large multiple sequence
alignments for each protein, the resulting species databases are not
independent. Orthologous proteins in each species will, by virtue of that fact,
produce profile HMMs with significant overlap.

Seed sequences for putative members of the Hawk family were selected
semi-arbitrarily for each species. Each seed was searched against the uniprot20
database using HHblits \cite{Remmert2011} (local alignment, two iterations).
Predicted secondary structure was added to each MSA/profile HMM using PSIPRED
\cite{Jones1999}. The resulting profile HMMs were then searched against the
relevant species-specific database using HHsearch (local alignment, single
iteration, no pre-filter) to generate a list of at most 500 putative paralogues
from each seed. In turn, each one of these sequences was subjected to the same
procedure, producing a large set of nodes and edges, with nodes representing
proteins and edges representing alignments between them, weighted by the rank of
the alignment.

The resulting graph was filtered by removing edges arising from alignments with
a length of less than 100 columns (accounting for the length of ~2 HEAT
repeats), an expect-value of greater than 0.01 (thus controlling the
false-discovery rate) or a true positive probability of less than 15\%. Edge
weights were then normalised according to the following formula, such that the
normalised rank $f(r)$ lies between 0.01 and 1.0, with 1.0 being the best
possible mean rank and 0.01 the worst.

\begin{displaymath}
    f(r) = \frac{1}{1 + \frac{99(r – r_{min} )}{r_{max} - r_{min}}}, 1 \leq r
    \leq 500
\end{displaymath}

At this stage, each edge has a direction, pointing from the protein used as a
query sequence to the returned paralogous protein. As such, a given pair of
nodes can be connected by either one edge or two - the former only being
possible if a protein appeared exclusively in the second round of searches and
was therefore not queried itself. In order to make the graph undirected, all
nodes with a degree of less than 2 were discarded and the remaining edges
between each pair of nodes combined and weighted by the geometric mean of
normalised alignment ranks. Since the geometric mean is always lower than the
arithmetic mean, this avoids giving too much weight to results from proteins
with very few significant alignments.

Finally, clustering was carried out using the mcl algorithm with an inflation
parameter I = 2.5 for all networks \cite{VanDongen2000}. Initial network
construction and parameter setting was performed on a fully-labelled \textit{S.
cerevisiae} network, but \textit{S. pombe} and \textit{H. sapiens} replicates
were performed on blinded graphs, with genes in each cluster only being revealed
after all filtering and cluster parameters had been fixed. GO term enrichment
analysis was carried out using the Cytoscape BiNGO app, with GO `Biological
Process' annotations \cite{Maere2005}. P-values were generated using the
hypergeometric test and corrected for false discovery rate using the
Benjamini-Hochberg method \cite{Benjamini1995,Maere2005}.

\subsubsection{Homology network permutation tests}
Assuming a null hypothesis under which alignment ranks contain no information
about the relative likelihood of two proteins being related, a single control
network was constructed for each species. This was generated from the observed
network by randomising the edge weights between each pair of nodes. This was
achieved by pre-filtering alignments as usual, but randomly assigning ranks.
These were then normalised and averaged as for the observed network. Each random
network was then clustered and each cluster tested for membership of Hawk
proteins; specifically we ask: does there exist a cluster in the random graph
containing exclusively those proteins from the largest Hawk cluster in the
observed graph? This process was repeated 106 times for each species, and the
resulting p-value calculated as the number of times the complete Hawk cluster
was seen, divided by the number of trials.

\subsubsection{Searching for lokiarchaeota HEAT repeat sequences}
13 Lokiarchaeota proteins containing HEAT repeats were downloaded from the
UniProt database; 9 on the basis of UniProt sequence annotations and an
additional 4 proteins, including 2 fragments, on the basis of HHsuite searches
and manual inspection. These sequences were searched against our human HHsuite
database, and the resulting human sequences searched back against the
lokiarchaeota database. A sub-graph was built using the same parameters as for
the main eukaryote networks, leaving exactly 10 archaeal proteins remaining
after quality control. The resulting set of edges was concatenated onto the
human network and re-clustered.

\subsubsection{Mapping of repeat domain boundaries}
Sequences from \textit{S. cerevisiae} hawks and clathrin adaptors were used to
generate multiple sequence alignments with HHblits. Multiple sequence alignments
were generated with the uniprot20\_2016\_02 database. These alignments were
subsequently passed to the HHRepID web server \newline
(https://toolkit.tuebingen.mpg.de/hhrepid). The threshold p-value for assigning
repeat domain families was kept at 0.01, and the threshold for suboptimal
self-alignments was set to 0.1, also the default. The number of HHblits
iterations was set to 0 since we had produced our own MSAs in the preceding
step. Repeat predictions were collected from the HHRepID results with alignment
stringencies between 0.0 and 0.3, depending on which value produced highest
confidence predictions.

\subsubsection{Structural alignments and conservation mapping}
Structures for human Pds5B and SA-2 were downloaded from the PDB (5HDT
\cite{Ouyang2016} and 4PJU \cite{Hara2014} respectively, 28.04.2016). Structures
were aligned in PyMol using TM-align \cite{PyMol2016,Zhang2005}, both globally
and locally by splitting SA-2 and Pds5B at residues L436 and Y462 respectively
and realigning each half. Conservation mapping was performed using multiple
sequence alignments generated as follows: For Pds5B and SA-2, 1000 metazoan
sequences for each were retrieved from the NCBI non-redundant sequence database
using blastp, then clustered to 90\% sequence identity with usearch
\cite{Altschul1990,Edgar2010}. The remaining sequences were then aligned in
forward and reverse directions with MAFFT, MUSCLE and GlProbs, with a final
composite MSA being generated with MergeAlign
\cite{Katoh2002,Edgar2004,Ye2015,Collingridge2012}. Finally, these were mapped
onto the PDB structures in Chimera \cite{Pettersen2004}.

\subsubsection{Analysis of putative Nse5 and Nse6 HEATS}
Specific searches for HEAT-containing Nse5 and Nse6 homologues were carried out
with the same parameters as for the main network – HHblits with 2 iterations to
generate profile HMMs, followed by HHsearch to find significant alignments in
the three main species datasets. Kre29 was used in place of Nse6 for S.
cerevisiae, and Slf2 for Human. Subsequent searches using HHblits/HHsearch were
carried out with more iterations for the HHblits step – this increases
sensitivity but at the cost of accuracy in determining relative rank of
alignments. Additional searches were performed in a wider variety of species
using the proteome datasets available on the HHSuite webserver. Next, HHRepID
\cite{Biegert2008} was used to try and detect repeats within Nse5-6 themselves
(as opposed to HEAT containing homologues). As before, human Slf2 was also
checked, as was Kre29. Iterations ranging from 3-8 were used to generate the
profile HMMs, thus spanning a wide range of sensitivities.

Finally, a literature search was performed to try and identify the published
evidence for the Nse5-6 HEAT annotations. On the basis of evidence for HEATs in
Nse6 presented by Pebernard et al. \cite{Pebernard2006}, we attempted to
replicate their finding using the structural prediction server 3D-PSSM, which is
now obsolete \cite{Kelley2000}. Following this, we used the Phyre2 web server
\cite{Kelley2015} with the Nse6 sequence (UniProt id - O13688) using default
settings.

\end{document}
